xolotl: An Intuitive and Approachable Neuron \& Network Simulator

### 2018-08-15 paper/paper.tex
General revisions of the manuscript

% new commands and shortcuts
\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}
\newcommand{\etal}{\textit{et al.}\xspace}
\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\etc}{\textit{etc.}\xspace}
\newcommand{\vs}{\textit{vs.}\xspace}
\newcommand{\viz}{\textit{viz.}\xspace}

@@

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed and power of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code promote reproducible and auditable research. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

@@

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python"  or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), and "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as "C" or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

This software is designed to be used from within MATLAB, a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make xolotl completely usable entirely from within MATLAB. Models created using this simulator appear in the MATLAB workspace as native objects, are are thus fully scriptable, and are fully compatible with the large library of toolboxes that MATLAB provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new C++ files on-the-fly. Finally, xolotl fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{A rich library of network components.} xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, Calcium level, and set of constituent components, representing either entire cells or parts of cells. Objects of type "conductance" represent populations of ion channels in a compartment that produce transmembrane currents. "synapse" objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type "mechanism" can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

\paragraph{Object-oriented.}
xolotl is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a xolotl model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the xolotl model tree (compartments, conductances, etc.) are bonafide MATLAB objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

@@

Our focus on our simulator's ease-of-use and speed imposed some limitations on its feature set.

@@

\paragraph{Inefficient tools for handling large networks.} xolotl was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, xolotl is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

@@

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in our simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the heirarchical nature of the model (Figure \ref{fig:figurehh}C).

Adding an injected current and calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function displays a voltage trace colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. This built-in feature reveals that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

Voltage clamping is an experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (\cite{dayanTheoreticalNeuroscience2001}). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, turrigianoSelectiveRegulationCurrent1995}).

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. xolotl makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

@@

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how xolotl can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, our simulator can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeller's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations. (\cite{deschutterConsumerGuideNeuronal1992}). xolotl makes it easy to bookmark configurations of a model and return to them at will. Internally, xolotl uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying C++ implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into xolotl leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

(Figure \ref{fig:figuresnapshot}) illustrates how these features can be used to explore the dynamics of the model presented in the previous section. First, the current state of the model is saved using the "snapshot" method into a state called ``initial''. In this state, the model exhibits periodic bursting activity due to a particular configuration of maximal conductance densities (Figure \ref{fig:figuremechanism}, orange).  On setting the maximal conductances of the Calcium-permissive channels to 0, the model switches to a tonic spiking activity (Figure \ref{fig:figuremechanism}, purple). Integrating the model for a longer duration allows the homeostatic control mechanism in the cell to restore the conductance profile and bursting activity to a state close to the original state (Figure \ref{fig:figuremechanism}, green). This state is saved using the name ``final''.

@@

Conductance-based neuron models are typically high dimensional and contain several parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

Our goal in designing xolotl was to create an easy-to-use neuron and network simulator that was fast enough and accurate enough for routine use by computational neuroscientists. In this section, we benchmarked the speed and accuracy of xolotl in simulating single and large numbers of Hodgkin-Huxley-like neuron models (as in Fig. \ref{fig:figurehh}) and bursting neuron models based on the bursting neurons in the lobster stomatogastric ganglion (STG) (\cite{prinzSimilarNetworkActivity2004}). For each type of neuron model, we compared our software to  NEURON (\cite{hinesNEURONSimulationEnvironment1997}), a high-performance and powerful neuron simulator specialized in simulating neurons with complex morphologies and DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), a general-purpose simulator that can solve coupled differential equations numerically. All simulators were run on the same hardware using fixed time-step solvers: xolotl used the Exponential Euler method (\cite{dayanTheoreticalNeuroscience2001}), NEURON used the  implicit Euler solver (\cite{hinesNEURONSimulationEnvironment1997}) and DynaSim used C-compiled 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance (\cite{sherfeyDynaSimMATLABToolbox2018}). We measured the speed of each simulator by dividing the time simulated for, by the time it took for the simulator to complete integration. For example, if a simulator could simulate 10 seconds of model data in 1 second, its speed would be 10X. We measured the speed of every simulator as a function of simulation time step, total length of simulation, and system size.

All three simulators were faster with larger time steps, since fewer iterations were needed (Fig. \ref{fig:figurebenchmark}A-B), and were approximately linear in the region tested. Our simulator compared favorably to NEURON and DynaSim in this task. We also measured the quality of the simulated output by comparing it to the simulated output at the smallest time step. Simulation error was measured using the LeMasson cost (\cite{lemassonIntroductionEquationSolving2000}), and was comparable amongst the three simulators (Fig. \ref{fig:figurebenchmark}C-D). Since our simulator sets up and runs the simulation in C++, it needs to transfer parameters and data to and from the underlying implementation. To measure the performance cost of this overhead, we repeated these benchmarks on all three simulators at a fixed time step of $0.1$ ms and varied the length of time simulated for. Speed increased with simulation duration up to a point, and then saturated, indicating a fixed performance cost to the overhead (Figure \ref{fig:figurebenchmark}E-F, black lines). Simulations using DynaSim, which also used a similar architecture and need to move data between C implementations and the MATLAB workspace, showed a similar increase in speed with simulation length (Figure \ref{fig:figurebenchmark}E-F, red lines). However, simulations using NEURON ran at a constant speed irrespective of simulation length, presumably due to differences in the underlying implementation (Figure \ref{fig:figurebenchmark}E-F, blue lines).

@@

Our simulator is a purely component-based simulator, and all equations need to be included in a C++ header file that specifies a object. Since our automatic type system binds MATLAB objects to the underlying C++ header files, objects can be inspected and parameters can be modified in the MATLAB workspace. To mitigate some of the drawbacks of the component-oriented paradigm, we have implemented an architecture that allows the user to access the underlying C++ code of any object by simply clicking on the object tree in the MATLAB command line. This feature removes the uncertainty inherent in other component-oriented simulators of the equations underlying each component, and allows the user to modify these equations if needed.

Another architectural choice in simulators is the syntax required for specifying models. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units in NEURON, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulator can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load of users using more than one simulator. In our software, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of our simulator.

@@

Since xolotl models are bonafide MATLAB objects, they are compatible with most of the powerful tools that exist within MATLAB. For example, it is possible to write simple scripts that run xolotl models in parallel using MATLAB's parallel processing toolbox, speeding up large simulations. xolotl models are also compatible with the global optimization toolbox in MATLAB, allowing parameters in xolotl models to be optimized, enabling the creation of toolboxes that efficiently tune parameters in neuron models to satisfy arbitrary constraints (\cite{Achard:2006ia, Krichmar:2014jj, Keren:2005ja, VanGeit:2007ea, Druckmann:2008kg}).

Care has been taken to reduce the amount of technical debt (\cite{suryanarayana2014refactoring}) associated with this project, with all parts of the simulator and dependencies written in a modular, objected oriented fashion. As a result, many of the key features and architectures of xolotl can be reused by others in their own applications. For example, cpplab, the automatic type system that binds C++ code to MATLAB objects, is independent of this simulator, and exists as a distinct repository that has been made freely available (https://github.com/sg-s/cpplab); and the framework for generating a GUI with sliders for each parameter that is hooked up to function callbacks also exists as an independent repository (https://github.com/sg-s/puppeteer) that can be easily integrated into other applications.

>>>>

% new commands and shortcuts
\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}
\newcommand{\etal}{\textit{et al.}\xspace}
\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\etc}{\textit{etc.}\xspace}
\newcommand{\vs}{\textit{vs.}\xspace}
\newcommand{\viz}{\textit{viz.}\xspace}
\newcommand{\cf}{\textit{cf.}\xspace}

@@

	Conductance-based models of neurons are used extensively in computational neuroscience, though working with these models continues to be a challenge due to their high dimensionality and large number of parameters. We present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed and power of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

@@

Two major approaches have dominated the design of neuron simulators. Component-based neuronal simulators, like NEURON (\cite{hinesNEURONSimulationEnvironment1997}), implemented in a fast compiled language like C, allow for the construction and simulation of neuron models using object-oriented paradigms.  Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python"  or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), and "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as "C" or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

xolotl is designed to be completely usable entirely from within MATLAB, a scientific programming language common among neuroscientists and engineers for research and pedagogy. Models created using this simulator appear in the MATLAB workspace as native objects, are are thus fully scriptable, and fully compatible with the large library of toolboxes that MATLAB provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new C++ files on-the-fly. Finally, xolotl fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{A rich library of network components.} xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components, representing either entire cells or parts of cells. Objects of type "conductance" represent populations of ion channels in a compartment that produce transmembrane currents. "synapse" objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type "mechanism" can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

\paragraph{Object-oriented.}
xolotl mirrors the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a xolotl model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the xolotl model tree (compartments, conductances, etc.) are genuine MATLAB objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain whole neurons or sections of membrane which contain populations of ion channels, making constructing, integrating, and thinking about these models easier.

@@

Our focus on ease-of-use and speed imposed some limitations on our simulator's feature set.

@@

\paragraph{Inefficient tools for handling large networks.} xolotl was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, xolotl is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like NEURON (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

@@

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in our simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the hierarchical nature of the model (Figure \ref{fig:figurehh}C).

Calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function displays a voltage trace colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. This built-in feature reveals that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

Voltage clamping is an experimental technique in which an amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (\cite{dayanTheoreticalNeuroscience2001}). Under this paradigm, the membrane voltage is ``clamped'' or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, turrigianoSelectiveRegulationCurrent1995}).

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Equation \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. xolotl makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

@@

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we show how xolotl can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, our simulator can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeler's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations (\cite{deschutterConsumerGuideNeuronal1992}). xolotl makes it easy to bookmark configurations of a model and return to them at will. Internally, xolotl uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying C++ implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into xolotl leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

Figure \ref{fig:figuresnapshot} illustrates how these features can be used to explore the dynamics of the model presented in the previous section. First, the current state of the model is saved using the "snapshot" method into a state called ``initial''. In this state, the model exhibits periodic bursting activity due to a particular configuration of maximal conductance densities (Figure \ref{fig:figuremechanism}, orange).  Upon setting the maximal conductances of the Calcium-permissive channels to 0, the model switches to a tonic spiking activity (Figure \ref{fig:figuremechanism}, purple). Integrating the model for a longer duration allows the homeostatic control mechanism in the cell to restore the conductance profile and bursting activity to a state close to the original state (Figure \ref{fig:figuremechanism}, green). This state is saved using the name ``final''.

@@

Conductance-based neuron models are typically high dimensional and contain several parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome; every step in this process involves ``mode'' changes: between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

Our goal in designing xolotl was to create an easy-to-use neuron and network simulator that was fast enough and accurate enough for routine use by computational neuroscientists. In this section, we benchmarked the speed and accuracy of xolotl in simulating single and large numbers of Hodgkin-Huxley-like neuron models (as in Fig. \ref{fig:figurehh}) and bursting neuron models based on the bursting neurons in the lobster stomatogastric ganglion (STG) (\cite{prinzSimilarNetworkActivity2004}). For each type of neuron model, we compared our software to  NEURON (\cite{hinesNEURONSimulationEnvironment1997}), a high-performance and powerful neuron simulator specialized in simulating neurons with complex morphologies, and DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), a general-purpose simulator that can solve coupled differential equations numerically. All simulators were run on the same hardware using fixed time-step solvers: xolotl used an exponential Euler method (\cite{dayanTheoreticalNeuroscience2001}), NEURON used an implicit Euler solver (\cite{hinesNEURONSimulationEnvironment1997}) and DynaSim used C-compiled 2\textsuperscript{nd}-order Runge-Kutta integration scheme as recommended for high-performance (\cite{sherfeyDynaSimMATLABToolbox2018}). We measured the speed of each simulator by dividing the time simulated, by the time it took for the simulator to complete integration. For example, if a simulator could simulate 10 seconds of model data in 1 second, its speed would be 10X. We measured the speed of every simulator as a function of simulation time step, total length of simulation, and system size.

All three simulators were faster with larger time steps, since fewer iterations were needed (Fig. \ref{fig:figurebenchmark}A-B). Speeds were approximately linear in the region tested. Our simulator compared favorably to NEURON and DynaSim in this task. We also measured the quality of the simulated output by comparing it to the simulated output at the smallest time step. Simulation error was measured using the LeMasson cost (\cite{lemassonIntroductionEquationSolving2000}), and was comparable amongst the three simulators (Fig. \ref{fig:figurebenchmark}C-D). Since our simulator sets up and runs the simulation in C++, it needs to transfer parameters and data to and from the underlying implementation. To measure the performance cost of this overhead, we repeated these benchmarks on all three simulators at a fixed time step of $0.1$ ms and varied the length of time simulated for. Speed increased with simulation duration up to a point, and then saturated, indicating a fixed performance cost to the overhead (Figure \ref{fig:figurebenchmark}E-F, black lines). Simulations using DynaSim, which also used a similar architecture and need to move data between C implementations and the MATLAB workspace, showed a similar increase in speed with simulation length (Figure \ref{fig:figurebenchmark}E-F, red lines). However, simulations using NEURON ran at a constant speed irrespective of simulation length, presumably due to differences in the underlying implementation (Figure \ref{fig:figurebenchmark}E-F, blue lines).

@@

Our simulator is a purely component-based simulator; all equations need to be included in a C++ header file that specifies a object. Since our automatic type system binds MATLAB objects to the underlying C++ header files, objects can be inspected and parameters can be modified in the MATLAB workspace. To mitigate some of the drawbacks of the component-oriented paradigm, we have implemented an architecture that allows the user to access the underlying C++ code of any object by simply clicking on the object tree in the MATLAB command line. This feature removes the uncertainty inherent in other component-oriented simulators of the equations underlying each component, and allows the user to modify these equations if needed.

The syntax required for specifying models is an architectural choice which varies between simulators. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units in NEURON, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulator can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load of users using more than one simulator. In our software, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of our simulator.

@@

Since xolotl models are genuine MATLAB objects, they are compatible with most of the powerful tools that exist within MATLAB. For example, it is possible to write simple scripts that run xolotl models in parallel using MATLAB's parallel processing toolbox, speeding up large simulations. xolotl models are also compatible with the global optimization toolbox in MATLAB, allowing parameters in xolotl models to be optimized, enabling the creation of toolboxes that efficiently tune parameters in neuron models to satisfy arbitrary constraints (\cite{Achard:2006ia, Krichmar:2014jj, Keren:2005ja, VanGeit:2007ea, Druckmann:2008kg}).

Care has been taken to reduce the amount of technical debt (\cite{suryanarayana2014refactoring}) associated with this project, with all parts of the simulator and dependencies written in a modular, objected oriented fashion. As a result, many of the key features and architectures of xolotl can be reused by others in their own applications. For example, "cpplab", the automatic type system that binds C++ code to MATLAB objects, is independent of this simulator, and exists as a distinct repository that has been made freely available (https://github.com/sg-s/cpplab); and the framework for generating a GUI with sliders for each parameter that is hooked up to function callbacks also exists as an independent repository (https://github.com/sg-s/puppeteer) that can be easily integrated into other applications.
